---
title: "McKinlab_Github_Monod_RpalFormate_2025_05.Rmd"
output: html_document
date: "2025-05-05"
author: "James 'Jake' B. McKinlay
---

## Overview
Monod model to test how the relationship between the formate oxidation rate and CO2-fixation rate affects culture growth.
Also used to test the effect of the initial CO2 concentration
Model assumes that reductant from formate oxidation (XH) can be discarded or stored
More information is available at: 

## Set working directory
```{r}
rm(list=ls()) # clears your working directory
getwd() # tells you your current working directory
setwd("/Users/jmckinla/Library/CloudStorage/OneDrive-IndianaUniversity/Lab_files_personal/Manuscripts_USE/Formate consumption") #~/GitHub/Task3") # sets your working directory
```

## Install packages
```{r}
#install.packages("deSolve") # run this first time
require("deSolve") # installs any dependencies
```

## Monod function
```{r}
mutfun <- function(times, y, parms) {
  with(as.list(c(y, parms)), { 
    uRp <- uMax*((CO2/(Kc+CO2))*(XH/(Kx+XH))*(For/(KF+For))) # Rp growth rate (h^-1)
    dCO2dt <- uRp*Rp/Yf - uRp*Rp/Yc # change in CO2 concentration (mM/h)
    dFdt <- -uRp*Rp/Yf # change in formate concentration (mM/h)
    dXdt <- uRp*Rp/Yf - uRp*Rp/Yx # change in reductant (e.g., NAD(P)H) concentration (mM/h)
    dRpdt <- uRp*Rp   # change in R. palustris cell density (cells/ml/h)
    return(list(c(dCO2dt, dFdt, dXdt, dRpdt))) 
  })
}
```
## Parameters and initial conditions
```{r}
# Parameters
parms <- c(
           uMax = 0.019, # R. palustris EIB max growth rate on formate alone w NH4Cl (h^-1); max growth rate for WT with formate + NaHCO3 is 0.0038 h^-1

           Yc = 4.8*10^7, # cells/umol CO2
#set lower boundary to be equal to formate yield because all CO2 has to come from formate and electrons will be limiting
# use these Yc values (*10^7) to alter the Formate oxidation rate : CO2 fixation rate by:
           #4.8 = 1.72X, 4.4 = 1.57X, 3.7 = 1.32 X, 2.8 = 1X
           #4.83*10^7 cell/umol is the parent R. palustris growth yield on NaHCO3 (CO2) with thiosulfate

           Yf = 2.8*10^7, # cells/umol formate (from R. palustris cultures with 20mM formate + 20 mM NaHCO3)

           Yx = 1.12*10^8, # cells per umol XH (NADPH)
#From McKinlay and Harwood 2010 PNAS, net of 14,270 umol XH used g cells; 625 mg/L/OD from McKinlay et al 2014 JBC -> 14270 umol/g x 0.625 g/L/OD = 8919 umol/L/OD ; 1 OD = 10^12 cells/L/OD; 10^12 cells/L/OD div by 8919 umol XH/L/OD = 1.12*10^8

           Kc = 0.067, # (mM); half saturation constant for CO2 (BRENDA; Witzel, F.; Goetze, J.; Ebenhoeh, O. Slow deactivation of ribulose 1,5-bisphosphate carboxylase/oxygenase elucidated by mathematical models (2010), FEBS J., 277, 931-950

           Kx = 0.0001, # (mM) set arbitrarily low to assume a saturating availability within the cell

           KF = 0.26) # (mM) half saturation constant for formate;  EC 1.17.1.9; Yoch, D.C.; Lindstrom, E.S. Nicotinamide adenine dinucleotide-dependent formate dehydrogenase from Rhodopseudomonas palustris (1969), Arch. Mikrobiol., 67, 182-188.

init <- c( 
          CO2 = 0.02, # (mM); use 0.0000001 for other conditions tested 
          For = 20, # (mM)
          XH = 0.1, # (mM)
          Rp = 13000000) # R. palustris (cells/ml)
          # Time range and time step
Tmax = 2400 # hours
TimeStep = 1 #hours
Time = seq(0, Tmax, by = TimeStep)
```

## Output
```{r}
out <-  ode(y = init, times = Time, func = mutfun, parms = parms, method = "lsode")
#head(out) # prints first 6 time points
tail(out) # prints last 6 time points
write.csv(out, file = "out.csv") # creates comma-delimited file that can be pasted into spreadsheet for graphing
```

## Plotting - not used to generate figures; use as a quick check of trends
```{r}
plot(out[,1], out[,5], xlab = "time (hr)", ylab = "Rp", type = "l", las = 1, 
     col = "purple", lwd = 4, ylim = c(0, 1*max(out[,5]))) # plot Rp biomass

par(new=T)
plot(out[,1], out[,2], xlab="", ylab ="", type = "l",las = 1,
     col = "black", ylim=c(0,1.2*max(out[,2])) )
axis(4, ylim=c(0,1.2*max(out[,2])), lwd=2, line=3.5)
mtext(4, text="CO2", line=5.5) # plot CO2 (mM)

par(new=T)
plot(out[,1], out[,3], xlab="", ylab ="",  type = "l",las = 1,
     col = "blue", ylim=c(0,1.3*max(out[,3])) )
axis(4, ylim=c(0,1.2*max(out[,3])), lwd = 4, line=4.5)
mtext(4, text="For", line=6.5) # plot For

par(new=T)
plot(out[,1], out[,4], xlab="", ylab ="",  type = "l",las = 1,
     col = "grey", ylim=c(0,1.3*max(out[,4])) )
axis(4, ylim=c(0,1.2*max(out[,4])), lwd = 4, line=4.5)
mtext(4, text="For", line=6.5) # plot XH

```